Nitric oxide feedback to ciliary photoreceptor cells gates a UV avoidance circuit

### 2023-12-19 Manuscript/Manuscript.Rmd
Update_response

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that *Platynereis* NOS belongs to an orthology group of bilaterian NOS sequences (Figure 1---figure supplement 1).
To characterise the expression pattern of *NOS*, we used in situ hybridization chain reaction (HCR) and transient transgenesis.
In two- and three-day-old larvae, we detected *NOS* expression in four cells (two of them weakly expressing) in the apical organ region (Figure 1E and Figure 1---figure supplement 2).
*NOS* was also expressed in the region of the visual eyes (adult eyes) and the pigmented eyespots (Figure 1---figure supplement 2).
The four apical organ cells, but not the eyes, were also labelled with a *NOS* reporter construct driving palmitoylated tdTomato (Figure 1F).
This reporter also revealed the axonal projections of these central NOS-expressing neurons.
The position and morphology of the four *NOS+* cells allowed us to identify the same four cells as four interneurons (INNOS) in our three-day-old whole-body *Platynereis* volume EM data [@williams2017; @veraszto2020] (Figure 1B,C).
In the synaptic connectome, the INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGWa interneurons, which are also cPRC targets (Figure 1C, H).
The projections of INNOS cells are segregated into input (dendritic) and output (axonal) compartments, with cPRC inputs occurring on the dendritic part in the dense neurosecretory plexus (Figure 1D).
Immunostaining revealed that the localization of NOS at the protein level is concentrated in the dendrites of the input region even within INNOS (Figure 1G).
INNOS output-synapses form in the more ventral projection region (Figure 1---figure supplement 3).
INRGWa neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1), which synapse on the cholinergic MC ciliomotor neuron and cells of the prototroch ciliary band (Figure 1C, H) [@veraszto2017].

@@

```{r echo=FALSE,fig.cap="**Figure 1 -- figure supplement 2. Expression of *NOS* in three-day-old *Platynereis* larva.** (**A**) HCR in situ for *NOS* (magenta), dorsal view. (**B**) HCR in situ for *NOS* (magenta), ventral view. Inset in (B) shows close-up of two *NOS*-positive cells (INNOS_dl and INNOS_vl) on the left side. Samples were also stained for DAPI (cyan) to label nuclei."}
knitr::include_graphics(path = "figures/Fig1_sup2.png")

@@

```{r echo=FALSE,fig.cap="**Figure 3 -- figure supplement 1. Generation and behavioural characterisation of *NOS* CRISPR knockouts.** (**A**) Top: The domain organisation of *Platynereis* NOS protein and exon/intron structure of the *NOS* gene. Bottom: The genomic locus of *NOS* with the CRISPR target site, wild-type and knockout (NOSΔ11, NOSΔ23) sequences and predicted protein sequences. The PAM sequence is shown in grey, stop codons in red. (**B**) Schematic diagram of the set-up of the behavioural experiment. (**C**) Swimming trajectories of wild type (WT, n=37) and *NOS* mutant (NOSΔ11/Δ11, n=18 and NOSΔ23/Δ23, n=8) two-day-old larvae. All trajectories start at 0 x and y position and time 0 corresponding to 10 sec before the onset of 395 nm stimulation from the side. (**D**)  Vertical displacement in 30 sec bins of wild type (WT, n=37) and mutant (NOSΔ11/Δ11, n=18 and NOSΔ23/Δ23, n=8) two-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. (**E**) Comparison of vertical position of wild type and mutant two-day-old larvae after 30 seconds of UV stimulation. One-way ANOVA with Dunnett’s multiple-comparison test are shown. The wild type (WT, n=52) and *NOS* mutant (*NOSΔ11/Δ11*, n=30 and *NOSΔ23/Δ23*, n=16) larvae. (**F**) Vertical position of batches of wild type (12 batches) and mutant (NOSΔ11/Δ11, 8 batches and NOSΔ23/Δ23, 3 batches) two-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. One-way ANOVA with Dunnett’s multiple-comparison test are shown. *P < 0.05. <br> Figure 3 -- figure suppmenet 1 -- source data 1-6. Source data for panels B-G."}
knitr::include_graphics(path = "figures/Fig3_sup1.png")

@@

```{r echo=FALSE,fig.cap="**Figure 3 -- figure supplement 2. Generation and behavioural characterisation of *NOS* CRISPR knockouts.** (**A,B**) Swimming speed of batches of wild type and *NOS* mutant two-day-old (WT, n=37, NOSΔ11/Δ11, n=18 and NOSΔ23/Δ23, n=8) (A) and three-day-old (WT, n=32, *NOSΔ11/Δ11*, n=26 and *NOSΔ23/Δ23*, n=47) (B) larvae under 395 nm UV stimulation. (**C,D**) Mean swimming speed of wild type and *NOS* mutant two-day-old (WT, n=37, NOSΔ11/Δ11, n=18 and NOSΔ23/Δ23, n=8) (C) and three-day-old (WT, n=32, *NOSΔ11/Δ11*, n=26 and *NOSΔ23/Δ23*, n=47) (D) larvae during 10 seconds before (dark) and 10 seconds after the start of 395 nm UV stimulation (sideUV). Data points for the same larva are joined by lines. One-tailed paired t-test; p-values (‘***’ <0.001) are shown. (**E,F**) Average ciliary beat frequency of wild type and mutant two-day-old (E) and three-day-old (F) single larvae during 10 seconds before (gray) and 10 seconds after the start of 395 nm UV stimulation (purple). N=6 larvae each. Data points for the same larva are joined by lines. One-tailed paired t-test; p-values (‘**’ <0.01 or ‘*’ <0.05) are shown. <br> Figure 3 -- figure suppmenet 2 -- source data 1-6. Source data for panels B-G."}
knitr::include_graphics(path = "figures/Fig3_sup2.png")

>>>>

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data.
Phylogenetic analysis of NOS proteins indicate that *Platynereis* NOS belongs to an orthology group of bilaterian NOS sequences (Figure 1---figure supplement 1).
To characterise the expression pattern of *NOS*, we used in situ hybridization chain reaction (HCR) and transient transgenesis.
In two- and three-day-old larvae, we detected *NOS* expression in four cells (two of them weakly expressing) in the apical organ region (Figure 1E and Figure 1---figure supplement 2).
*NOS* was also expressed in the region of the visual eyes (adult eyes) and the pigmented eyespots (Figure 1---figure supplement 2).
The four apical organ cells, but not the eyes, were also labelled with a *NOS* reporter construct driving palmitoylated tdTomato (Figure 1F).
This reporter also revealed the axonal projections of these central NOS-expressing neurons.
The position and morphology of the four *NOS+* cells allowed us to identify the same four cells as four interneurons (INNOS) in our three-day-old whole-body *Platynereis* volume EM data [@williams2017; @veraszto2020] (Figure 1B,C).
In the synaptic connectome, the INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGWa interneurons (cholinergic neurons that strongly express RGW neuropeptide), which are also cPRC targets (Figure 1C, H).
The projections of INNOS cells are segregated into input (dendritic) and output (axonal) compartments, with cPRC inputs occurring on the dendritic part in the dense neurosecretory plexus (Figure 1D).
Immunostaining revealed that the localization of NOS at the protein level is concentrated in the dendrites of the input region even within INNOS (Figure 1G).
INNOS output-synapses form in the more ventral projection region (Figure 1---figure supplement 3).
INRGWa neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1), which synapse on the cholinergic MC ciliomotor neuron and cells of the prototroch ciliary band (Figure 1C, H) [@veraszto2017].

@@

```{r echo=FALSE,fig.cap="**Figure 1 -- figure supplement 2. Expression of *NOS* in three-day-old *Platynereis* larva.** (**A**) HCR in situ for *NOS* (magenta), dorsal view. (**B**) HCR in situ for *NOS* (magenta), ventral view. Inset in (B) shows close-up of two *NOS*-positive cells (INNOS_dl and INNOS_vl) on the left side. Samples were also stained for DAPI (cyan) to label nuclei.(**C**) Expression of the *NOS* gene detected by in situ HCR (magenta) in a three-day-old larva (anterior view). Antibody staining for acetylated α-tubulin (acTub: green) highlights cPRC cilia and the neuropil."}
knitr::include_graphics(path = "figures/Fig1_sup2.png")

@@

```{r echo=FALSE,fig.cap="**Figure 3 -- figure supplement 1. Generation and behavioural characterisation of *NOS* CRISPR knockouts.** (**A**) Top: The domain organisation of *Platynereis* NOS protein and exon/intron structure of the *NOS* gene. Bottom: The genomic locus of *NOS* with the CRISPR target site, wild-type and knockout (NOSΔ11, NOSΔ23) sequences and predicted protein sequences. The PAM sequence is shown in grey, stop codons in red. **(B, C)** Immunostaining for NOS antibody in NOS mutant (B, NOSΔ11/Δ11; C, NOSΔ23/Δ23) larvae. Acetylated α-tubulin staining (green) highlights the neuropil and cPRC cilia. Anterior view of a three-day-old larva."}
knitr::include_graphics(path = "figures/Fig3_sup1.png")

@@

```{r echo=FALSE,fig.cap="**Figure 3 -- figure supplement 2. Generation and behavioural characterisation of *NOS* CRISPR knockouts.** (**A**) Schematic diagram of the set-up of the behavioural experiment. (**B**) Swimming trajectories of wild type (WT, n=37) and *NOS* mutant (NOSΔ11/Δ11, n=18 and NOSΔ23/Δ23, n=8) two-day-old larvae. All trajectories start at 0 x and y position and time 0 corresponding to 10 sec before the onset of 395 nm stimulation from the side. (**C**)  Vertical displacement in 30 sec bins of wild type (WT, n=37) and mutant (NOSΔ11/Δ11, n=18 and NOSΔ23/Δ23, n=8) two-day-old larvae stimulated with 395 nm light from the side, 488 nm light from the top and 395 nm light from the top. (**D**) Comparison of vertical position of wild type and mutant two-day-old larvae after 30 seconds of UV stimulation. One-way ANOVA with Dunnett’s multiple-comparison test are shown. The wild type (WT, n=52) and *NOS* mutant (*NOSΔ11/Δ11*, n=30 and *NOSΔ23/Δ23*, n=16) larvae. (**E**) Vertical position of batches of wild type (12 batches) and mutant (NOSΔ11/Δ11, 8 batches and NOSΔ23/Δ23, 3 batches) two-day-old larvae over time under 395 nm UV stimulation. The starting position of each larval trajectory was set to 0. One-way ANOVA with Dunnett’s multiple-comparison test are shown. *P < 0.05. <br> Figure 3 -- figure suppmenet 1 -- source data 1-6. Source data for panels B-G."}
knitr::include_graphics(path = "figures/Fig3_sup2.png")

```{r echo=FALSE,fig.cap="**Figure 3 -- figure supplement 3. Generation and behavioural characterisation of *NOS* CRISPR knockouts.** (**A,B**) Swimming speed of batches of wild type and *NOS* mutant two-day-old (WT, n=37, NOSΔ11/Δ11, n=18 and NOSΔ23/Δ23, n=8) (A) and three-day-old (WT, n=32, *NOSΔ11/Δ11*, n=26 and *NOSΔ23/Δ23*, n=47) (B) larvae under 395 nm UV stimulation. (**C,D**) Mean swimming speed of wild type and *NOS* mutant two-day-old (WT, n=37, NOSΔ11/Δ11, n=18 and NOSΔ23/Δ23, n=8) (C) and three-day-old (WT, n=32, *NOSΔ11/Δ11*, n=26 and *NOSΔ23/Δ23*, n=47) (D) larvae during 10 seconds before (dark) and 10 seconds after the start of 395 nm UV stimulation (sideUV). Data points for the same larva are joined by lines. One-tailed paired t-test; p-values (‘***’ <0.001) are shown. (**E,F**) Average ciliary beat frequency of wild type and mutant two-day-old (E) and three-day-old (F) single larvae during 10 seconds before (gray) and 10 seconds after the start of 395 nm UV stimulation (purple). N=6 larvae each. Data points for the same larva are joined by lines. One-tailed paired t-test; p-values (‘**’ <0.01 or ‘*’ <0.05) are shown. <br> Figure 3 -- figure suppmenet 2 -- source data 1-6. Source data for panels B-G."}
knitr::include_graphics(path = "figures/Fig3_sup3.png")
